Overrelaxed Sinkhorn--Knopp Algorithm for Regularized Optimal Transport

### 2017-10-20 paper.tex
Relecture Nicolas

Efficient parallel computations can be considered \cite{cuturi13} and one can even reach real-time computation for large scale problem for certain class of cost matrices $c$ \cite{Solomon2015}. 
For small values of the parameter $\epsilon$, numerical issues can arise and a stabilization of the algorithm is necessary \cite{2016arXiv160705816C}.
The convergence of the process can nevertheless be very slow when $\epsilon$ is small.

@@

The resulting algorithm, which is proved to be convergent by Theorem~\ref{thm:algo}, is written in pseudo-code in Algorithm \ref{SOR}. It uses the function $\Theta$ defined implicitly in \eqref{eq:Theta}, which in practice is approximated with a few iterations of Newton's method on the function $\omega \mapsto \varphi_\omega(\min u)$ which is decreasing is as can be seen on Figure~\ref{phi_omega}. With the choice $\theta_0=1$, one recovers exactly the original SK algorithm.is \vspace{-0.1cm}
\begin{algorithm}
\caption{Overrelaxed SK algorithm }
\label{SOR}
\begin{algorithmic}
\REQUIRE $\mu^1\in \IR^{n_1}$, $\mu^2\in \IR^{n_2}$, $c\in \IR^{n_1\times n_2}_+$
\STATE Set $a=\mathbf{1}_{n_1}$, $b=\mathbf{1}_{n_2}$, $\gamma^0=e^{-c/\epsilon}$, $\theta_0\in[1;2)$ and $\eta>0$
\WHILE {$||a\otimes \gamma^0 b -  \mu_1||>\eta$}
\STATE $\tilde a=\mu_1\oslash (\gamma^0 b)$, 
\STATE $\omega=\Theta(a\oslash\tilde a)$
%\IF{ $\langle \mu^1,\varphi_\omega(a\oslash\tilde a) \rangle>0$}
\STATE  $a=a^{1-\omega}\otimes \tilde a^\omega$
%\ELSE
%\STATE $a=\tilde a$,\COMMENT{SK iteration}
%\ENDIF
\STATE $\tilde b=\mu_2\oslash (^t\gamma^0  a)$
\STATE $\omega=\Theta(b\oslash\tilde b)$
%\IF{ $\langle \mu^2,\varphi_\omega(b\oslash\tilde b) \rangle>0$}
\STATE  $b=b^{1-\omega}\otimes \tilde b^\omega$%\COMMENT{Overrelaxed SK iteration}
%\ELSE
%\STATE $b=\tilde b$,\COMMENT{SK iteration}
%\ENDIF
\ENDWHILE
 \RETURN $\gamma=\diag(a)\gamma^0\diag(b)$
\end{algorithmic}\vspace*{-0.1cm}
\end{algorithm}

\section{Acceleration of local convergence rate}
\label{section:local}
In order to justify the acceleration of convergence that is observed in practice, we now study the local convergence rate of the overrelaxed algorithm, which follows from the classical convergence analysis of the linear SOR method. Our result involves the second largest eigenvalue of the matrix
\begin{equation}\label{eq:M1}
M_1= \diag(1\oslash \mu^1) \, \gamma^* \, \diag(1\oslash \mu^2)\, ^t\gamma^*
\end{equation}
where $\gamma^*$ is the solution to the regularized OT problem (the largest eigenvalue is $1$, associated to the eigenvector $\mathbf{1}$). We denote this second largest eigenvalue by $1-\eta$, which satisfies $\eta>0$ (see \cite{knight2008sinkhorn}).

\begin{proposition}\label{prop:local}
The SK algorithm converges locally at a linear rate $1-\eta$. For the optimal choice of extrapolation parameter $\theta^* = 2/(1+\sqrt{\eta})$, the overrelaxed projection algorithm converges locally linearly at a rate $(1-\sqrt{\eta})/(1+\sqrt{\eta})$. More generally, the local convergence of the overrelaxed algorithm is guaranteed as long as $\theta \in {]0,2[}$ and the linear rate is given on Figure~\ref{fig:eigtransform} as a function of $1-\eta$ and $\theta$.
\end{proposition}

@@

To switch from these dual convergence results to primal convergence results, remark that $\gamma^\ell \to \gamma^*$ implies $\KL(\gamma^\ell,\gamma^0)\to \KL(\gamma^*,\gamma^0)$ which in turn implies $E(\alpha^\ell,\beta^\ell) \to \max E$ so invoking the partial strict concavity of $E$, $\pi_S(\alpha^\ell,\beta^\ell)\to (\alpha^*,\beta^*)$. The converse implication is direct so it holds $[\pi_S(\alpha^\ell,\beta^\ell)\to (\alpha^*,\beta^*)] \Leftrightarrow [\gamma^\ell \to \gamma^*]$. We conclude by noting that the fact that $\pi_S(\alpha^\ell, \beta^\ell)$ converges at a linear rate implies the same rate on $\gamma^\ell$, thanks to the relationship between the iterates.
%Using standard arguments, we have that the local convergence rate of the SOR procedure on the dual objective $E$ is that of the SOR procedure applied to the quadratic form defined by its Hessian and the linear rate of \emph{semi-convergence} to the projection on the set of minimizers. 
\end{proof}

\begin{figure}
\centering
\includegraphics[scale=0.65]{images/eigen_transform}
\caption{\label{fig:eigtransform} Local linear rate of convergence of the overrelaxed algorithm as a function of $1-\eta$, the local convergence rate of SK algorithm and $\theta$ the overrelaxation parameter. (plain curve) the original rate is recovered for $\theta=1$. (dashed curve) optimal overrelaxation parameter $\theta^*$.}
\end{figure}

>>>>

Efficient parallel computations can be considered \cite{cuturi13} and one can almost reach real-time computation for large scale problem for certain class of cost matrices $c$  allowing the use of seprable convolutions \cite{Solomon2015}. 
For small values of the parameter $\epsilon$, numerical issues can arise and a stabilization of the algorithm is necessary \cite{2016arXiv160705816C}.
The convergence of the process can nevertheless be very slow when $\epsilon$ is small.

@@

The resulting algorithm, which is proved to be convergent by Theorem~\ref{thm:algo}, is written in pseudo-code in Algorithm \ref{SOR}. It uses the function $\Theta$ defined implicitly in \eqref{eq:Theta}, which in practice is approximated with a few iterations of Newton's method on the function $\omega \mapsto \varphi_\omega(\min u)$ which is decreasing is as can be seen on Figure~\ref{phi_omega}. With the choice $\theta_0=1$, one recovers exactly the original SK algorithm.is 
\begin{algorithm}
\caption{Overrelaxed SK algorithm }
\label{SOR}
\begin{algorithmic}
\REQUIRE $\mu^1\in \IR^{n_1}$, $\mu^2\in \IR^{n_2}$, $c\in \IR^{n_1\times n_2}_+$
\STATE Set $a=\mathbf{1}_{n_1}$, $b=\mathbf{1}_{n_2}$, $\gamma^0=e^{-c/\epsilon}$, $\theta_0\in[1;2)$ and $\eta>0$
\WHILE {$||a\otimes \gamma^0 b -  \mu_1||>\eta$}
\STATE $\tilde a=\mu_1\oslash (\gamma^0 b)$, 
\STATE $\omega=\Theta(a\oslash\tilde a)$
%\IF{ $\langle \mu^1,\varphi_\omega(a\oslash\tilde a) \rangle>0$}
\STATE  $a=a^{1-\omega}\otimes \tilde a^\omega$
%\ELSE
%\STATE $a=\tilde a$,\COMMENT{SK iteration}
%\ENDIF
\STATE $\tilde b=\mu_2\oslash (^t\gamma^0  a)$
\STATE $\omega=\Theta(b\oslash\tilde b)$
%\IF{ $\langle \mu^2,\varphi_\omega(b\oslash\tilde b) \rangle>0$}
\STATE  $b=b^{1-\omega}\otimes \tilde b^\omega$%\COMMENT{Overrelaxed SK iteration}
%\ELSE
%\STATE $b=\tilde b$,\COMMENT{SK iteration}
%\ENDIF
\ENDWHILE
 \RETURN $\gamma=\diag(a)\gamma^0\diag(b)$
\end{algorithmic}
\end{algorithm}

\section{Acceleration of local convergence rate}
\label{section:local}
In order to justify the acceleration of convergence that is observed in practice, we now study the local convergence rate of the overrelaxed algorithm, which follows from the classical convergence analysis of the linear SOR method. Our result involves the second largest eigenvalue of the matrix
\begin{equation}\label{eq:M1}
M_1= \diag(1\oslash \mu^1) \, \gamma^* \, \diag(1\oslash \mu^2)\, ^t\gamma^*
\end{equation}
where $\gamma^*$ is the solution to the regularized OT problem (the largest eigenvalue is $1$, associated to the eigenvector $\mathbf{1}$). We denote the second largest eigenvalue by $1-\eta$, it satisfies $\eta>0$  \cite{knight2008sinkhorn}.

\begin{proposition}\label{prop:local}
The SK algorithm converges locally at a linear rate $1-\eta$. For the optimal choice of extrapolation parameter $\theta^* = 2/(1+\sqrt{\eta})$, the overrelaxed projection algorithm converges locally linearly at a rate $(1-\sqrt{\eta})/(1+\sqrt{\eta})$. The local convergence of the overrelaxed algorithm is guaranteed for $\theta \in {]0,2[}$ and the linear rate is given on Figure~\ref{fig:eigtransform} as a function of $1-\eta$ and $\theta$.
\end{proposition}

@@

To switch from these dual convergence results to primal convergence results, remark that $\gamma^\ell \to \gamma^*$ implies $\KL(\gamma^\ell,\gamma^0)\to \KL(\gamma^*,\gamma^0)$ which in turn implies $E(\alpha^\ell,\beta^\ell) \to \max E$ so invoking the partial strict concavity of $E$, $\pi_S(\alpha^\ell,\beta^\ell)\to (\alpha^*,\beta^*)$. The converse implication is direct so it holds $[\pi_S(\alpha^\ell,\beta^\ell)\to (\alpha^*,\beta^*)] \Leftrightarrow [\gamma^\ell \to \gamma^*]$. We conclude by noting the fact that $\pi_S(\alpha^\ell, \beta^\ell)$ converges at a linear rate implies the same rate on $\gamma^\ell$, thanks to the relationship between the iterates.
%Using standard arguments, we have that the local convergence rate of the SOR procedure on the dual objective $E$ is that of the SOR procedure applied to the quadratic form defined by its Hessian and the linear rate of \emph{semi-convergence} to the projection on the set of minimizers. 
\end{proof}

\begin{figure}
\centering
\includegraphics[scale=0.8]{images/eigen_transform}
\caption{\label{fig:eigtransform} Local linear rate of convergence of the overrelaxed algorithm as a function of $1-\eta$, the local convergence rate of SK algorithm and $\theta$ the overrelaxation parameter. (plain curve) the original rate is recovered for $\theta=1$. (dashed curve) optimal overrelaxation parameter $\theta^*$.}
\end{figure}
